### 2021-05-14 newvariant.tex
already cut down 873 words

%% 6635 words
%% 6040 words

@@

Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from inference data under the assumption that the true generation-interval distribution of the wild type is known.
We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Supplementary Text):  
(1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
(2) the variant has a shorter mean generation interval (\fref{Rtbias}D--F); and
(3) the variant has a longer mean generation interval (\fref{Rtbias}G--I).
Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Supplementary Text) given the possibility that behavioral changes may affect transmission before and after interventions take place.
We reach similar conclusions if we use a step function instead (Supplementary Figure S1).

@@

When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}B).
However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
In particular, if the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).

@@

While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to reduce $\rv$ below 0 (or, equivalently, $\Rv$ below 1).

@@

We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading disease variant.
Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.
Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
these biases may be assessed by considering whether relative strength appears to vary systematically with underlying epidemiological dynamics.
Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions:
if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).

As new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
Our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions about epidemiological conditions and generation-interval distributions.
However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.

@@

We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data:
when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
In our simulations, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
however, this assumption only holds under strength-like interventions, which are insensitive to age of infection, but not under speed-like interventions. 
As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution;
therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.

>>>>

%% 6635 words
%% 5762 words

@@

Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from inference data under the assumption that the true generation-interval distribution of the wild type is known.
We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Supplementary Text):  
(1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
(2) the variant has a shorter mean generation interval (\fref{Rtbias}D--F); and
(3) the variant has a longer mean generation interval (\fref{Rtbias}G--I).
Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
In order to simulate introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ to decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
Previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, but we use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Supplementary Text) given the possibility that behavioral changes may affect transmission before and after interventions take place.
We reach similar conclusions if we use a step function instead (Supplementary Figure S1).

@@

When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}B).
However, when the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
If the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate $\Rv(t)$ during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).

@@

While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant determine the spread and conditions for control of the variant over the long term.
In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to reduce $\rv$ below 0 (or, equivalently, $\Rv$ below 1).

@@

We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading disease variant.
Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.
Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
these biases may be assessed by considering whether relative strength appears to vary systematically with underlying epidemiological dynamics.
Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
% if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).

As new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.

@@

We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data:
when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
In our simulations, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution;
therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.
